A Mechanised Library for Generic (Co)Programming

### 2024-02-23 paper/paper.tex
Writing for section 5

\subsection{Sized Types}

Another approach to certifying termination of nonstructurally recursive
functions is the use of \emph{sized types}. Sized types were introduced in
\cite{HughesPS96} as a way to track/verify various properties of recursive
programs, including productivity and termination. The core idea of sized types
is that types express bounds on the sizes of recursive data structures. This
approach has been used to express nonstructurally recursive algorithms in Agda:
\cite{CopelloTB14} used sized types in a straightforward formulation of
mergesort. Such an approach has also been used in MiniAgda~\cite{abs-1202-3496}.

>>>>

\subsection{Termination of Nonstructural Recursion}

Termination and productivity are non-trivial properties of programs, and various
methods have been proposed for checking that these properties hold. A common
approach is guardedness checking~\cite{10.5555/646535.695850}, which is a
\emph{syntactic} check that definitions avoid the introduction of
non-normalizable terms. However, some desirable definitions will not be accepted
by such checks.

In particular, the problem of \emph{nonstructural recursion} is
well-studied~\cite{BOVE_KRAUSS_SOZEAU_2016}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
Take, for instance, division by iterated subtraction, which is inherently
non-structurally recursive since it involves recursion based on the result of a
subtraction. There is a nonstandard implementation of divivion found on Coq's
standard library, which involves a four-argument function that effectively
combines subtraction and division. Similarly, the mergesort in Coq's standard
library uses an ``explicit stack of pending merges'' in order to avoid issues
with nonstructural definitions. There is a major downside, however; as noted by
Abreu et al., the result is ``barely recognizable as a form of
mergesort''~\cite{AbreuDHJMS23}.

\subsubsection{Sized Types}

Another approach to certifying termination of nonstructurally recursive
functions is the use of \emph{sized types}. Sized types were introduced by
Hughes et al.\ as a way to track/verify various properties of recursive
programs, including productivity and termination~\cite{HughesPS96}. The core
idea of sized types is that types express bounds on the sizes of recursive data
structures. With this approach, algorithms are implemented in the standard
functional way, following the divide-and-conquer pattern of splitting, recurring,
and merging, with the addition of the data types being indexed by a static
approximation of the relative \emph{size}. An advantage of this approach is
that it allows for code to be written in a natural way.

This approach has been used to express nonstructurally recursive
algorithms in Agda: Copello et al.\ used sized types in a straightforward
formulation of mergesort~\cite{CopelloTB14}. Such an approach has also been used
in MiniAgda~\cite{abs-1202-3496}. A proposal exists to add sized types to
Coq as well~\cite{CHAN_2023}.

\subsubsection{Divide and Conquer Recursion}

Abreu et al.\ propose\ldots~\cite{AbreuDHJMS23}. Their approach has similarities to what is presented here.

They differ in requiring the assumption of the functional extensionality axiom.
Additionally, they are concerned with the erasure of termination proofs but have
little discussion of what the resulting extracted code \emph{looks like} (that
is, whether the extracted OCaml code resembles the natural formulation of the
recursive function).
\mvol{We should find out for sure if they use \texttt{obj.magic}.}
